What is an Ancestral Recombination Graph?

### 2022-02-08 paper.tex

\section*{ARG inference methods}
Using the terminology developed here to classify ARGs, we review methods developed
to \emph{infer} ARGs from sequencing data. We focus our attention on some recent methods and discuss the
properties of the ARGs inferred.

The problem of reconstructing ARGs for samples of recombining sequences has been of significant interest
essentially since the ARG was first defined. Early methods focussed on finding the most
\emph{parsimonious} ARGs, i.e.\ those minimising the number of
posited recombination events in the history of the sample \citep{hein1990reconstructing, song2003parsimonious,
lyngso2005minimum}. However, reconstructing a most parsimonious ARG for a given data set
is NP-hard \citep{wang2001perfect}, and in order to allow for larger input datasets,
subsequent parsimony-based methods have resorted to various relaxations and heuristics \citep{hein1993heuristic,
song2005efficient, wu2008association, camara2016inference, thao2019hybrid, ignatieva2021kwarg}.
The scalability of these methods varies substantially based on the heuristic in use
and the underlying data structure, with the most efficient tools limited to analysing hundreds of sequences.
A number of heuristic methods have also been developed which do not (explicitly) focus on reconstructing
the most parsimonious ARGs, but rather aim for computational efficiency
while seeking to achieve good accuracy of the inferred genealogies
\citep{minichiello2006mapping, parida2008estimating, wu2011new, mirzaei2017rent}; some of these methods \citep{kelleher2019inferring, speidel2019method, zhang2021biobank} can be practically
used on megabase scale data and tens to hundreds of thousands of samples under human-like
parameters.
% AI: just to note, these papers do not all agree in their definition of an ARG.
% I guess this is one of the points of the present paper, but should we mention this fact?
% Should we also mention that most of these infer just the topology, but some also the times?
% How much more description of the parsimony methods would be good? Maybe a mention of them splitting into
% "back in time" and "across local trees" types?

An alternative approach is to treat the ancestry as a latent parameter to be averaged out
by Monte Carlo methods, based either on importance sampling
\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian, fallon2013acg}.
These methods invariably operated on a representation of the ``little ARG", typically
generated until the grand MRCA. They are extremely computationally expensive,
and applicable to at most hundreds of samples consisting of tens of kilobases with
human-like parameters.

>>>>

\section*{ARG inference methods}
Using the terminology developed here to classify ARGs, we now review methods developed
to explicitly \emph{infer} ARGs from sequencing data.

\subsection*{Parsimony and heuristics}
The problem of reconstructing ARGs for samples of recombining sequences has been of significant
 interest essentially since the ARG was first defined. Early methods focussed on finding the most 
 \emph{parsimonious} ARGs, i.e.\ those minimising the number of posited recombination events in 
 the history of the sample \citep{hein1990reconstructing}. Two main approaches to explicitly reconstructing 
 parsimonious ARGs then emerged: building the ARG either \emph{backwards-in-time} or \emph{along-the-genome}. 
 The former approach, introduced by \citet{lyngso2005minimum} and implemented in the program 
 \texttt{Beagle}, starts with the input data matrix and reduces it through row and column operations 
 corresponding to coalescence, mutation and recombination events, until the empty matrix is reached; 
 the corresponding sequence of events can then be used to construct an ARG from the bottom up. The 
 latter approach, introduced by \citet{song2003parsimonious} and first implemented in the program 
 \texttt{RecMinPath} \citep{song2005constructing}, starts with an initial local tree at some starting 
 position, and refines the tree moving left and right along the genome, changing the topology of the 
 tree when necessary through \emph{subtree prune and regraft} (SPR) operations (corresponding to 
 recombination events).

Reconstructing a most parsimonious ARG for a given data set is NP-hard \citep{wang2001perfect}, 
and in order to allow for larger input datasets, subsequent parsimony-based methods have resorted 
to various relaxations and heuristics. Those utilising variants of the backwards-in-time approach 
include the programs \texttt{SHRUB} \citep{song2005efficient}, \texttt{GAMARG} \citep{thao2019hybrid}, 
and \texttt{KwARG} \citep{ignatieva2021kwarg}, and the method described by \citet{wu2008association}. 
Tools constructing genealogies along-the-genome include \texttt{RecPars} \citep{hein1993heuristic}, 
\texttt{RENT} \citep{wu2011new} and its sequel \texttt{RENT+} \citep{mirzaei2017rent}. Other approaches 
include \texttt{TARGet} \citep{camara2016inference}, which implements a method based on topological 
data analysis. The scalability of these methods varies substantially based on the heuristics in use 
and the underlying data structure, with the most efficient tools limited to analysing hundreds of 
sequences.

A number of heuristic methods have also been developed which do not (explicitly) focus on reconstructing 
the most parsimonious ARGs, but rather aim for computational efficiency while seeking to achieve good 
accuracy of the inferred genealogies. Such methods include \texttt{Margarita} \citep{minichiello2006mapping} 
and the approach of \citet{parida2008estimating}; the more recently developed \texttt{Relate} 
\citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} use a combination of both 
backwards-in-time and along-the-genome approaches, and can be practically used on megabase scale data 
and tens to hundreds of thousands of samples under human-like parameters.
Another class of methods has recently emerged that use a \emph{threading} approach, constructing ARGs 
by adding sequences to the graph one-by-one. Originally introduced by \citet{rasmussen2014genome} as a 
way of generating MCMC proposals for ARGs, a similar idea was leveraged in \texttt{ARGneedle} 
\citep{zhang2021biobank} to reconstruct ARGs for biobank-scale data.
% AI: just to note, these papers do not all agree in their definition of an ARG.
% I guess this is one of the points of the present paper, but should we mention this fact?
% Should we also mention that most of these infer just the topology, but some also the times?

\subsection*{Model-based methods}
An alternative approach is to treat the ancestry as a latent parameter to be averaged out
by Monte Carlo methods, based either on importance sampling
\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian, fallon2013acg}.
These methods invariably operated on a representation of the ``little ARG", typically
generated until the grand MRCA. They are extremely computationally expensive,
and applicable to at most hundreds of samples consisting of tens of kilobases with
human-like parameters.
